---
title: "3-bayes-hlm"
output: html_document
date: "2023-05-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file I standardized the data, applied Bayes HLM to a few other metabolites individually, ran some posterior diagnostics, and applied (sterilization status x size) + (sex x size) as a group.

l# Standardize Data

```{r}
# Standardize metabolite abundances
metabolites_data_t_std <- data.frame(elab_id=metabolites_data_t$elab_id, scale(metabolites_data_t[,-1]))

# Merge std data
meta_traits_data_std <- merge(dog_info, metabolites_data_t_std, by='elab_id', all.x=TRUE)

# Factorize data
meta_traits_data_std$size_at_DOC <- as.factor(meta_traits_data_std$size_at_DOC)

# Standardize age (scale is creating diff dimensions...)
meta_traits_data_std$age_std <- scale(meta_traits_data_std$age)[,1]
```

# 1) Group by Size

## Run CmdStan

```{r}
hier_reparam_model_v3 <- cmdstan_model("stan-models/hlm_metabolite_reparam_v3.stan")

hier_reparam_fit_v3_std <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_at_DOC),
                                                        group=as.integer(meta_traits_data_std$size_at_DOC),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$N.Ac.Phenylalanine),
                                  refresh=0, show_messages=FALSE)

```

## Run Stan

```{r}
hier_reparam_fit2_v3_std <- stan(file="hlm_metabolite_reparam_v3.stan",
                                 data=list(N=length(meta_traits_data_std$elab_id),
                                           K=nlevels(meta_traits_data_std$size_at_DOC),
                                           group=as.integer(meta_traits_data_std$size_at_DOC),
                                           age = meta_traits_data_std$age_std,
                                           metabolite = meta_traits_data_std$N.Ac.Phenylalanine))
```

```{r}
test_sum <- as.data.frame(summary(hier_reparam_fit2_v3_std), reset_index=TRUE)
#test_sum %>% filter(grepl("^beta[1]\\[", variable))$summary.mean
test_sum %>% rownames_to_column(var = "Variable")
```


```{r}
plot(hier_reparam_fit2_v3_std, pars=c("beta0[5]", "beta0[4]","beta0[3]","beta0[2]", "beta0[1]"))
plot(hier_reparam_fit2_v3_std, pars=c("beta1[5]", "beta1[4]","beta1[3]","beta1[2]", "beta1[1]"))
```



## Posterior Interval

```{r}
hier_reparam_fit_v3_std$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() +
  scale_y_discrete(labels = c("Giant", "Large", "Medium", "Standard", "Toy")) +
  ggtitle("95% C.I. of N.Ac.Phenylalanine Abundance in Different Dog Sizes") +
  xlab("Metabolite Abundance") +
  ylab("Size")
```

Large diffrence bewteen Large to Giant due to lack of data? 

## Shrinkage Plot

```{r}
# posterior mean
post_means_v3 <- hier_reparam_fit_v3_std$summary() %>% 
  select(variable, mean) %>% 
  filter(grepl("^beta[01]\\[", variable))
#mle_v3 <- coef(summary(lm(N.Ac.Phenylalanine ~ age_centered + size_at_DOC, data = meta_traits_data)))[2:3, 1]

# MLE 
grouped_data <- split(meta_traits_data_std, as.integer(meta_traits_data_std$size_at_DOC))
lm_output <- lapply(grouped_data, function(data) lm(N.Ac.Phenylalanine ~ age_std, data = data))
mle_coef_v3 <- lapply(lm_output, coef)
mle_coef_v3 <- data.frame(do.call(rbind, mle_coef_v3)) # turn MLE list into a dataframe

# data frame of PM and MLE of intercepts
pm_mle_b1 <- data.frame(
  size = c(1:5),
  mle_b1 = mle_coef_v3$age_std,
  pm_b1 = (post_means_v3 %>% filter(grepl("^beta[1]\\[", variable)) %>% select(mean))
)

global_mean_b1 <- hier_reparam_fit_v3_std$summary() %>% 
  select(variable, mean) %>% 
  filter(grepl("^mu_beta1", variable))


# plot
shrinkage_plot(pm_mle_b1$mle_b1, pm_mle_b1$mean, shrink_point=global_mean_b1$mean) 
```


## Fitted Values
(generated quantities)

```{r}
hier_reparam_fit_v3_std$summary() %>% 
  filter(grepl("^beta[0]\\[", variable))
```

```{r}
# Extract beta's
beta0_v3 <- hier_reparam_fit_v3_std$summary() %>% 
  filter(grepl("^beta[0]\\[", variable))
beta1_v3 <- hier_reparam_fit_v3_std$summary() %>% 
  filter(grepl("^beta[1]\\[", variable))
beta0_v3
beta1_v3
```
```{r}
ggplot(meta_traits_data_std, aes(x = age_std, y = N.Ac.Phenylalanine)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE) + 
  facet_wrap(~ size_at_DOC) +
  
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Giant"), aes(slope=0.5367173, intercept=0.029281847), size=1, color="red") +
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Giant"), aes(slope=0.35238570	, intercept=-0.07349504), size=1, color="pink") + # lower
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Giant"), aes(slope=0.7469767, intercept=0.17568825), size=1, color="pink") + # upper 
  geom_ribbon(data=subset(meta_traits_data_std, size_at_DOC=="Giant"), aes(x=age_std, ymin=-0.07349504+(0.35238570*age_std), ymax=0.17568825+(0.7469767*age_std)), fill="pink") +
  
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Large"), aes(slope=0.2099390, intercept=-0.024043044), size=1, color="red") +
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Medium"), aes(slope=0.3935481, intercept=-0.027091217), size=1, color="red") +
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Standard"), aes(slope=0.4057697, intercept=-0.025849529), size=1, color="red") +
  geom_abline(data=subset(meta_traits_data_std, size_at_DOC=="Toy"), aes(slope=0.4519261, intercept=-0.004283106), size=1, color="red") 
```

Makes sense for larger dogs to have more intense slope because they age quicker.

# 2) Group by Size x Sterilization Status

```{r}
# Create new column for size x sterilization 
meta_traits_data_std$size_x_sterilization <- paste(meta_traits_data_std$size_at_DOC, meta_traits_data_std$sterilization_status, sep = "-")
meta_traits_data_std$size_x_sterilization <- as.factor(meta_traits_data_std$size_x_sterilization)
# Even numbers are 1 (sterile), odd numbers are 0
# 10 is Toy and 1 is Giant
```

```{r}
hier_reparam_fit_v3_size_x_steril <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_x_sterilization),
                                                        group=as.integer(meta_traits_data_std$size_x_sterilization),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$N.Ac.Phenylalanine),
                                  refresh=0, show_messages=FALSE)
```

## Posterior Intervals

```{r, fig.width=10cm, fig.height=5cm}
library(gridExtra)
par(mfrow=c(1,2))

no_sterile_b1 <- hier_reparam_fit_v3_size_x_steril$summary() %>% 
  filter(grepl("^beta[1]\\[[13579]\\]", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() +
  scale_x_continuous(limits=c(-0.25,1))

sterile_b1 <- hier_reparam_fit_v3_size_x_steril$summary() %>% 
  filter(grepl("^beta[1]\\[[02468]\\]", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() +
  scale_x_continuous(limits=c(-0.25,1))

grid.arrange(no_sterile_b1, sterile_b1, ncol=2)
```

```{r, fig.width=8, fig.height=4}
category_order <- c("beta1[1]", "beta1[2]", "beta1[3]", "beta1[4]", "beta1[5]", "beta1[6]", "beta1[7]", "beta1[8]", "beta1[9]", "beta1[10]")

df_filtered <- hier_reparam_fit_v3_size_x_steril$summary() %>% 
  filter(grepl("^beta1\\[(?:[1-9]|10)\\]", variable)) %>% 
  mutate(integer = as.numeric(sub("^beta1\\[(?:[1-9]|10)\\]", "\\1", variable))) %>%
  arrange(integer) 

df_filtered$variable <- factor(df_filtered$variable, levels=category_order)
levels(df_filtered$variable) <- c("Yes x Giant", "No x Giant", "Yes x Large", "No x Large", "Yes x Medium", "No x Medium", "Yes x Standard", "No x Standard", "Yes x Toy", "No x Toy")

df_filtered$sterilization <- ifelse(grepl("Yes", df_filtered$variable), "sterile", "not sterile")

# levels(df_filtered$variable) <- c("Giant", "Large", "Medium", "Standard", "Toy", "Giant", "Large", "Medium", "Standard", "Toy")

ggplot(df_filtered, aes(y=variable, x=mean, xmin=q5, xmax=q95, color=sterilization)) +
  geom_pointinterval() +
  labs(x="Sterilization x Size", y="Metabolite Abundance",title = "95% C.I. of N.Ac.Phenylalanine Slope in Different Sterilization x Size of Dogs") 

```


```{r}
hier_reparam_fit_v3_size_x_steril$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() 
```

Sterile (even) has more certainty.

```{r}
table(meta_traits_data_std$sterilization_status)
```

## Significance of Sterility 

```{r}
# See if the difference between sterile vs nonsterile is significant 

# To measure the difference wouldn't it need to be paired? (beta11-beta12)
as.data.frame(hier_reparam_fit_v3_size_x_steril$draws())
#get_drawset(hier_reparam_fit_v3_size_x_steril, pars='mu_beta0')
hier_reparam_fit_v3_size_x_steril$draws(format="df") %>% colnames()
```

## Shrinkage Plot

```{r}
# posterior mean
post_means_sterile_v3 <- hier_reparam_fit_v3_size_x_steril$summary() %>% 
  select(variable, mean) %>% 
  filter(grepl("^beta[01]\\[", variable))
#mle_v3 <- coef(summary(lm(N.Ac.Phenylalanine ~ age_centered + size_at_DOC, data = meta_traits_data)))[2:3, 1]

# MLE 
grouped_data <- split(meta_traits_data_std, as.integer(meta_traits_data_std$size_x_sterilization))
lm_output <- lapply(grouped_data, function(data) lm(N.Ac.Phenylalanine ~ age_std, data = data))
mle_coef_sterile_v3 <- lapply(lm_output, coef)
mle_coef_sterile_v3 <- data.frame(do.call(rbind, mle_coef_sterile_v3)) # turn MLE list into a dataframe

# data frame of PM and MLE of intercepts
pm_mle_sterile_b1 <- data.frame(
  size = c(1:10),
  mle_b1 = mle_coef_sterile_v3$age_std,
  pm_b1 = (post_means_v3 %>% filter(grepl("^beta[1]\\[", variable)) %>% select(mean))
)
pm_mle_sterile_b1

# global mean
global_mean_sterile_b1 <- hier_reparam_fit_v3_size_x_steril$summary() %>% 
  select(variable, mean) %>% 
  filter(grepl("^mu_beta1", variable))
```

```{r}
shrinkage_plot(pm_mle_sterile_b1$mle_b1, pm_mle_sterile_b1$mean, shrink_point=global_mean_sterile_b1$mean)
```


# Is the robust difference between 1 and 2 common in other metabolites?

Glycine shows negative correlation between age and abundance.

```{r}
# fit on another metabolite
hier_reparam_fit_v3_std_glycine <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_at_DOC),
                                                        group=as.integer(meta_traits_data_std$size_at_DOC),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$Glycine),
                                  refresh=0, show_messages=FALSE)
```

```{r}
b1_post_interval <- function(fit) {
  fit$summary() %>% 
    filter(grepl("^beta[]\\[", variable)) %>% 
    ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
   geom_pointinterval() 
}

b1_post_interval(hier_reparam_fit_v3_std_glycine)
```

```{r}
ggplot(meta_traits_data_std, aes(x = age_std, y = Glycine)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE) + 
  facet_wrap(~ size_at_DOC)
```


```{r}
# try another one
hier_reparam_fit_v3_std_carnitine <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_at_DOC),
                                                        group=as.integer(meta_traits_data_std$size_at_DOC),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$Carnitine),
                                  refresh=0, show_messages=FALSE)
```
```{r}
b1_post_interval(hier_reparam_fit_v3_std_carnitine)
```

No obvious pattern. Always more uncertain for Giant dogs (because less data). 

```{r}
ggplot(meta_traits_data_std, aes(x = age_std, y = Carnitine)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE) + 
  facet_wrap(~ size_at_DOC)
```


# 3) Group By Sex x Size

```{r}
# Create new column for size x sex 
meta_traits_data_std$size_x_sex <- paste(meta_traits_data_std$size_at_DOC, meta_traits_data_std$sex, sep = "-")
meta_traits_data_std$size_x_sex <- as.factor(meta_traits_data_std$size_x_sex)
# Even numbers are 1 (sterile), odd numbers are 0
# 10 is Toy and 1 is Giant

hier_reparam_fit_v3_size_x_sex <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_x_sex),
                                                        group=as.integer(meta_traits_data_std$size_x_sex),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$N.Ac.Phenylalanine),
                                  refresh=0, show_messages=FALSE)
```
```{r}
b1_post_interval(hier_reparam_fit_v3_size_x_sex)
# Male = 1 (even), Female = 0 (odd)
# 10 = toy, 1=giant
```

Has that curve pattern where from Large to Giant there is a jump

# 4) Group by Metabolite 

```{r}
# Rearrange data so we're isoalte for giant size

# Select elab_id, age, size, and all metabolites
#head(meta_traits_data_std)
#meta_traits_data_std[,27:164]

meta_traits_data_std_long <- meta_traits_data_std %>%
  pivot_longer(cols = colnames(meta_traits_data_std[,27:163]), names_to = "Metabolite", values_to = "Value")
# Add age_std back
meta_traits_data_std_long$age_std <- scale(meta_traits_data_std_long$age)[,1]
head(meta_traits_data_std_long)

# Filter just giant dogs
meta_traits_data_std_long_giant <- meta_traits_data_std_long %>% filter(size_at_DOC == "Giant")
dim(meta_traits_data_std_long_giant)
# Make Metabolite categorical
meta_traits_data_std_long_giant$Metabolite <- as.factor(meta_traits_data_std_long_giant$Metabolite)

# Filter just toy dogs
meta_traits_data_std_long_toy <- meta_traits_data_std_long %>% filter(size_at_DOC == "Toy")
dim(meta_traits_data_std_long_toy)
# Make Metabolite categorical
meta_traits_data_std_long_toy$Metabolite <- as.factor(meta_traits_data_std_long_toy$Metabolite)
```

```{r}
hier_reparam_fit_v3_giant <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std_long_giant$elab_id), 
                                                        K=nlevels(meta_traits_data_std_long_giant$Metabolite),
                                                        group=as.integer(meta_traits_data_std_long_giant$Metabolite),
                                                        age = meta_traits_data_std_long_giant$age_std,
                                                        metabolite = meta_traits_data_std_long_giant$Value),
                                  refresh=0, show_messages=FALSE)
```

```{r}
giant_sum <- hier_reparam_fit_v3_giant$summary()
giant_sum$not_intersecting_0 <- !((giant_sum$q5 <= 0) & (giant_sum$q95 >= 0)) 
giant_sum <- giant_sum[order(giant_sum$mean), ]
giant_sum_ordered <- giant_sum %>% arrange(mean)
giant <- giant_sum %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=reorder(variable,mean), x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval(aes(color = not_intersecting_0)) +
  ggtitle(label="95% C.I. of Metabolite Abundance in Giant Dogs") +
  ylab("Slope of Different Metabolites") +
  xlab("Metabolite Abundance") +
  scale_x_continuous(limit=c(-0.75,0.75)) 
giant
#ggplotly(giant, tooltip=c("variable"))
```


```{r}
hier_reparam_fit_v3_toy <- hier_reparam_model_v3$sample(data=list(N=length(meta_traits_data_std_long_toy$elab_id), 
                                                        K=nlevels(meta_traits_data_std_long_toy$Metabolite),
                                                        group=as.integer(meta_traits_data_std_long_toy$Metabolite),
                                                        age = meta_traits_data_std_long_toy$age_std,
                                                        metabolite = meta_traits_data_std_long_toy$Value),
                                  refresh=0, show_messages=FALSE)
```

```{r}
toy_sum <- hier_reparam_fit_v3_toy$summary()
toy_sum$not_intersecting_0 <- !((toy_sum$q5 <= 0) & (toy_sum$q95 >= 0)) 
toy <- toy_sum %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=reorder(variable, mean), x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval(aes(color = not_intersecting_0)) +
  ggtitle(label="95% C.I. of Metabolite Abundance in Toy Dogs") +
  ylab("Slope of Different Metabolites") +
  xlab("Metabolite Abundance") +
  scale_x_continuous(limit=c(-0.75,0.75)) 
  # theme(axis.text.y = element_text(vjust = 10))  
toy_sum
```

```{r}
# replace each beta1 variable with number that maps to it

## create a dataframe mapping of integer to metabolite name
integers <- c(as.integer(meta_traits_data_std_long_toy$Metabolite))
metas <- c(meta_traits_data_std_long_toy$Metabolite)
meta_map <- data.frame(int=integers, metabolites=metas)

toy_beta1 <- toy_sum %>% 
  filter(grepl("^beta[1]\\[", variable))

replace_with_mapping <- function(value) {
  index <- as.numeric(gsub("beta1\\[(\\d+)\\]", "\\1", value))
  mapped_value <- meta_map$metabolites[meta_map$int == index]
  if (length(mapped_value) > 0) {
    return(mapped_value)
  } else {
    return(NA)
  }
}

# Apply the function to replace values
toy_beta1$new_var <- sapply(toy_beta1$variable, replace_with_mapping)
toy_beta1
```

```{r, fig.width=10, fig.height=13}
toy_beta1$variable <- factor(toy_beta1$variable)
levels(toy_beta1$variable) <- meta_traits_data_std_long_toy$Metabolite

ggplot(toy_beta1, aes(y=reorder(variable, mean), x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval(aes(color = not_intersecting_0)) +
  ggtitle(label="95% C.I. of Metabolite Abundance in Toy Dogs") +
  ylab("Slope of Different Metabolites") +
  xlab("Metabolite Abundance") +
  scale_x_continuous(limit=c(-0.75,0.75)) 
```
```{r, fig.width=10, fig.height=13}
giant_beta1 <- giant_sum %>% 
  filter(grepl("^beta[1]\\[", variable))
giant_beta1$not_intersecting_0 <- !((giant_beta1$q5 <= 0) & (giant_beta1$q95 >= 0)) 

giant_beta1$variable <- factor(giant_beta1$variable)
levels(giant_beta1$variable) <- meta_traits_data_std_long_toy$Metabolite

ggplot(giant_beta1, aes(y=reorder(variable, mean), x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval(aes(color = not_intersecting_0)) +
  ggtitle(label="95% C.I. of Metabolite Abundance in Giant Dogs") +
  ylab("Slope of Different Metabolites") +
  xlab("Metabolite Abundance") +
  scale_x_continuous(limit=c(-0.75,0.75)) 
```


```{r}
table(meta_traits_data_std$size_at_DOC)
```
```{r}
hier_reparam_fit_v3_giant$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  filter(mean>0.5)
```
```{r}
hier_reparam_fit_v3_giant$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) 
```

```{r}
all_meta <- meta_traits_data_std_long_toy$Metabolite
meta_number <- as.integer(meta_traits_data_std_long_toy$Metabolite)
cbind.data.frame(all_meta,meta_number)
```

```{r}
data.frame(table(meta_traits_data_std$breed_1))
```

