---
title: "4-all-metabolites"
output: html_document
date: "2023-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
meta_traits_data_std
```

```{r}
meta <- "X1.3.Methylhistidine"
# meta_traits_data_std[, meta] 
# colnames(meta_traits_data_std)
metas <- c(colnames(metabolites_data_t_std[, 2:ncol(metabolites_data_t_std)]))
```

```{r}
# Need to be CAREFUL of naming in metas (make sure exact name exists)
extract_beta1 <- function(df, metas, stan_model) {
  # Create empty data frame to store values
  beta1_df <- data.frame(
    metabolite = character(),
    type = numeric(),
    beta1 = numeric(),
    stringsAsFactors=F
  )
  
  for (meta in metas) {
    # Fit stan model on data
    stan_fit <- stan_model$sample(data=list(N=length(df$elab_id), 
                                                        K=nlevels(df$size_at_DOC),
                                                        group=as.integer(df$size_at_DOC),
                                                        age = df$age_std,
                                                        metabolite = df[, meta]),
                                  refresh=0, show_messages=FALSE)
    
    # Find beta1 values for all 5 groups
    meta_beta1 <- stan_fit$summary() %>% 
      filter(grepl("^beta[1]\\[", variable))
    
    # if age and meta is scaled, beta0 should be 1
    
    # Add information to dataframe
    for (i in 1:nrow(meta_beta1)) {
      row <- meta_beta1[i,]
      # return (row)
      # names(row) <- names(beta1_df) # preserve column names
      meta_row <- c(meta, row$variable, row$mean) # create new row
      # print (meta_row)
      beta1_df[nrow(beta1_df)+1,] <- meta_row
      # beta1_df <- rbind(beta1_df, meta_row) # add to df
    }
  }
  return (beta1_df)
}
```

```{r}
# Split by positive and negative!!!
meta_temp <- c('N.Ac.Phenylalanine', 'N.Ac.Tryptophan', 'Acetylcarnitine', 'Glycine', 'Deoxycarnitine', 'X4.Guanidinobutanoate')
# meta_temp <- c('N.Ac.Phenylalanine', 'N.Ac.Tryptophan')
beta1_df <- extract_beta1(meta_traits_data_std, meta_temp, hier_reparam_model_v3)
head(beta1_df)
```

# Functions

```{r}
plot_beta1 <- function(beta1_df) {
  ggplot(beta1_df, aes(x=type, y=beta1, color=metabolite)) +
    geom_line(aes(group=metabolite)) +
    theme(legend.position="none") +
    geom_hline(yintercept=0, linetype='dashed') +
    scale_y_continuous(limits = c(-0.4, 0.4)) 
}
```

# Test on Original Model

## Positively Correlated

```{r}
# Find all positive correlated metabolites
pos_meta <- which(age_diff$est > 0) # age_diff is all metas w/ significant relationship between age and meta
pos_meta_names <- age_diff[pos_meta, ]$Met

# Calculate beta1
pos_beta1_df <- extract_beta1(meta_traits_data_std, pos_meta_names, hier_reparam_model_v3)
```

```{r}
age_diff
```

## Signifcance by Group

```{r}
toy_size <- meta_traits_data_std %>% filter(size_at_DOC == "Toy")
standard_size <- meta_traits_data_std %>% filter(size_at_DOC == "Standard")
medium_size <- meta_traits_data_std %>% filter(size_at_DOC == "Medium")
large_size <- meta_traits_data_std %>% filter(size_at_DOC == "Large")
giant_size <- meta_traits_data_std %>% filter(size_at_DOC == "Giant")

test_age_significance <- function(df) {
  model <- lm(metabolite_abundance ~ age, data = df)
  p_value <- summary(model)$coefficients[2, "Pr(>|t|)"]
  return(p_value < 0.05)
}

toy_significant <- sapply(split(toy_size, toy_size$metabolite_abundance), test_age_significance)

```


```{r}
pos_beta1_df$type <- as.factor(pos_beta1_df$type)
pos_beta1_df$beta1 <- as.numeric(pos_beta1_df$beta1)
pos_bayes <- plot_beta1(pos_beta1_df) 
pos_bayes_int <- ggplotly(pos_bayes, tooltip=c("type", "beta1", "metabolite"))
```

Group 2 dipping is somewhat evident here.

```{r}
# Take all metas in this list, calculate the slopes of each size group and plot it
get_freq_slopes <- function(metas, data) {
  
  beta1_df <- data.frame(
    metabolite = character(),
    type = numeric(),
    beta1 = numeric(),
    stringsAsFactors=F
  )
  
  for (meta in metas) {
    # Calculate slope for each size group
    for (size in levels(data$size_at_DOC)) {
      size_data <- data %>% filter(size_at_DOC==size)
      slope <- summary(lm(size_data[, meta]~size_data$age))$coefficients[2]
      # add row
      meta_row <- c(meta, size, numeric(slope)) # create new row
      beta1_df[nrow(beta1_df)+1,] <- meta_row
    }
  }
  return (beta1_df)
}
```

```{r}
size_data_test <- meta_traits_data_std %>% filter(size_at_DOC=="Giant")
test_sum <- summary(lm(size_data_test[, "X3HBA"]~size_data_test$age))
test_sum
```


```{r, fig.height=5, fig.width=15}
pos_freq_df <- get_freq_slopes(pos_meta_names, meta_traits_data_std)
pos_freq_df$beta1 <- as.numeric(pos_freq_df$beta1)
pos_freq <- plot_beta1(pos_freq_df)

grid.arrange(pos_freq + ggtitle("Frequentist: Slopes of Significant Positively Correlated Metabolites"), pos_bayes + ggtitle("Bayes: Slopes of Significant Positively Correlated Metabolites"), nrow = 1)
```

```{r}
pos_freq_df
```


```{r}
# pos_bayes_int
meta_traits_data_std$n.Formylmethionine
```


## Negatively Correlated

```{r}
# Find all negatively correlated metabolites
neg_meta <- which(age_diff$est < 0) # age_diff is all metas w/ significant relationship between age and meta
neg_meta_names <- age_diff[neg_meta, ]$Met

# Calculate beta1
neg_beta1_df <- extract_beta1(meta_traits_data_std,neg_meta_names,hier_reparam_model_v3)
```

```{r, fig.height=5, fig.width=15}
neg_beta1_df$type <- as.factor(neg_beta1_df$type)
neg_beta1_df$beta1 <- as.numeric(neg_beta1_df$beta1)
neg_bayes <- plot_beta1(neg_beta1_df)
neg_bayes_int <- ggplotly(neg_bayes, tooltip=c("type", "beta1", "metabolite"))

neg_freq_df <- get_freq_slopes(neg_meta_names, meta_traits_data_std)
neg_freq_df$beta1 <- as.numeric(neg_freq_df$beta1)
neg_freq <- plot_beta1(neg_freq_df)
grid.arrange(neg_freq + ggtitle("Frequentist: Slopes of Significant Negatively Correlated Metabolites"), neg_bayes + ggtitle("Bayes: Slopes of Significant Negatively Correlated Metabolites"), nrow = 1)
```

Can see a very subtle pattern which goes against the hypothesis (smaller dogs would be less intense beta).

# Change Prior on Shrinkage Variance

```{r}
hier_reparam_model_v3.1 <- cmdstan_model("stan-models/hlm_metabolite_reparam_v3.1.stan")
```


```{r}
# check model
hier_reparam_fit_v3.1 <- hier_reparam_model_v3.1$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_at_DOC),
                                                        group=as.integer(meta_traits_data_std$size_at_DOC),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$N.Ac.Phenylalanine),
                                  refresh=0, show_messages=FALSE)

hier_reparam_fit_v3.1$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() 
```


```{r}
pos_beta1_df3 <- extract_beta1(meta_traits_data_std, pos_meta_names, hier_reparam_model_v3.1)
```

```{r}
pos_beta1_df3$type <- as.factor(pos_beta1_df3$type)
pos_beta1_df3$beta1 <- as.numeric(pos_beta1_df3$beta1)
pos_bayes3 <- plot_beta1(pos_beta1_df3) 
pos_bayes_int3 <- ggplotly(pos_bayes3, tooltip=c("type", "beta1", "metabolite"))
pos_bayes3
```

## Shrink Variance More

```{r}
hier_reparam_model_v3.2 <- cmdstan_model("stan-models/hlm_metabolite_reparam_v3.2.stan")

# check model
hier_reparam_fit_v3.2 <- hier_reparam_model_v3.2$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_at_DOC),
                                                        group=as.integer(meta_traits_data_std$size_at_DOC),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$N.Ac.Phenylalanine),
                                  refresh=0, show_messages=FALSE)

hier_reparam_fit_v3.2$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() 
```

```{r}
hier_reparam_model_v3.3 <- cmdstan_model("stan-models/hlm_meta_reparam_v3.3.stan")

# check model
hier_reparam_fit_v3.3 <- hier_reparam_model_v3.3$sample(data=list(N=length(meta_traits_data_std$elab_id), 
                                                        K=nlevels(meta_traits_data_std$size_at_DOC),
                                                        group=as.integer(meta_traits_data_std$size_at_DOC),
                                                        age = meta_traits_data_std$age_std,
                                                        metabolite = meta_traits_data_std$N.Ac.Phenylalanine),
                                  refresh=0, show_messages=FALSE)

hier_reparam_fit_v3.3$summary() %>% 
  filter(grepl("^beta[1]\\[", variable)) %>% 
  ggplot(aes(y=variable, x=mean, xmin=q5, xmax=q95)) +
  geom_pointinterval() 
```

```{r}
pos_beta1_df3.3 <- extract_beta1(meta_traits_data_std, pos_meta_names, hier_reparam_model_v3.3)

pos_beta1_df3.3$type <- as.factor(pos_beta1_df3.3$type)
pos_beta1_df3.3$beta1 <- as.numeric(pos_beta1_df3.3$beta1)
pos_bayes3.3 <- plot_beta1(pos_beta1_df3.3) 
#pos_bayes_int3.3 <- ggplotly(pos_bayes3.2, tooltip=c("type", "beta1", "metabolite"))
pos_bayes3.3
```


```{r}
pos_beta1_df3.2 <- extract_beta1(meta_traits_data_std, pos_meta_names, hier_reparam_model_v3.2)

pos_beta1_df3.2$type <- as.factor(pos_beta1_df3.2$type)
pos_beta1_df3.2$beta1 <- as.numeric(pos_beta1_df3.2$beta1)
pos_bayes3.2 <- plot_beta1(pos_beta1_df3.2)

pos_bayes_int3.2 <- ggplotly(pos_bayes3.2, tooltip=c("type", "beta1", "metabolite"))
pos_bayes3.2
```

```{r}
pos_bayes3 <- plot_beta1(pos_beta1_df3) 
pos_bayes3 <- plot_beta1(pos_beta1_df3) 
pos_bayes3.2 <- plot_beta1(pos_beta1_df3.2)
pos_bayes3.3 <- plot_beta1(pos_beta1_df3.3) 
```



```{r, fig.width=25, fig.height=5}
grid.arrange(pos_bayes3.3 + ggtitle("Bayes (Shrink Cauchy(0,10))"), pos_bayes + ggtitle("Bayes (Shrink Cauchy(0,1))"), pos_bayes3 + ggtitle("Bayes (Shrink N(0,0.1))"), pos_bayes3.2 + ggtitle("Bayes (Shrink N(0,0.01))"), nrow = 1) 
# shrink more towards beta1 global mean = slope of other sizes (shrinkage is seen in horizontal line)
# giant has small n so it's now closer to the slopes of the other groups 
```

```{r}
neg_beta1_df3 <- extract_beta1(meta_traits_data_std, neg_meta_names, hier_reparam_model_v3.1)
neg_beta1_df3.2 <- extract_beta1(meta_traits_data_std, neg_meta_names, hier_reparam_model_v3.2)

neg_beta1_df3$type <- as.factor(neg_beta1_df3$type)
neg_beta1_df3$beta1 <- as.numeric(neg_beta1_df3$beta1)
neg_bayes3 <- plot_beta1(neg_beta1_df3) 

neg_beta1_df3.2$type <- as.factor(neg_beta1_df3.2$type)
neg_beta1_df3.2$beta1 <- as.numeric(neg_beta1_df3.2$beta1)
neg_bayes3.2 <- plot_beta1(neg_beta1_df3.2) 
```

```{r, fig.width=25, fig.height=5}
grid.arrange(neg_freq + ggtitle("Frequentist: Slopes of Significant Negatively Correlated Metabolites"), neg_bayes + ggtitle("Bayes (Shrink Cauchy(0,1)): Slopes of Significant Negatively Correlated Metabolites"), neg_bayes3 + ggtitle("Bayes (Shrink N(0,0.1)): Slopes of Significant Negatively Correlated Metabolites"), neg_bayes3.2 + ggtitle("Bayes (Shrink N(0,0.01)): Slopes of Significant Negatively Correlated Metabolites"), nrow = 1) 
```



# Test on Model with Inputs for Age, Sex, and Sterlilization

Rewrite `extract_beta` to match new model

```{r}
# Need to be CAREFUL of naming in metas (make sure exact name exists)
extract_beta1_v2 <- function(df, metas, stan_model) {
  # Create empty data frame to store values
  beta1_df <- data.frame(
    metabolite = character(),
    type = numeric(),
    beta1 = numeric(),
    stringsAsFactors=F
  )
  
  for (meta in metas) {
    # Fit stan model on data
    stan_fit <- stan_model$sample(data=list(N=length(df$elab_id), 
                                                        K=nlevels(df$size_at_DOC),
                                                        group=as.integer(df$size_at_DOC),
                                                        age = df$age_std,
                                                        sex = df$sex,
                                                        sterilization = df$sterilization_status,
                                                        metabolite = df[, meta]),
                                  refresh=0, show_messages=FALSE)
    
    # Find beta1 values for all 5 groups
    meta_beta1 <- stan_fit$summary() %>% 
      filter(grepl("^beta[1]\\[", variable))
    
    # Add information to dataframe
    for (i in 1:nrow(meta_beta1)) {
      row <- meta_beta1[i,]
      # names(row) <- names(beta1_df) # preserve column names
      meta_row <- c(meta, row$variable, row$mean) # create new row
      beta1_df[nrow(beta1_df)+1,] <- meta_row
      # beta1_df <- rbind(beta1_df, meta_row) # add to df
    }
  }
  return (beta1_df)
}
```


## Positively Correlated

```{r}
# Calculate beta1
pos_beta1_df2 <- extract_beta1_v2(meta_traits_data_std, pos_meta_names,hier_reparam_model_v5)
pos_beta1_df2$type <- as.factor(pos_beta1_df2$type)
pos_beta1_df2$beta1 <- as.numeric(pos_beta1_df2$beta1)
```

```{r}
plot_beta1(pos_beta1_df2)
plot_beta1(pos_beta1_df)
```

## Negatively Correlated

```{r}
neg_beta1_df2 <- extract_beta1_v2(meta_traits_data_std,neg_meta_names,hier_reparam_model_v5)
```

```{r}
neg_beta1_df2$type <- as.factor(neg_beta1_df2$type)
neg_beta1_df2$beta1 <- as.numeric(neg_beta1_df2$beta1)
plot_beta1(neg_beta1_df2)
```

